A Type-Theoretical Wide-Coverage Computational Grammar for Swedish

### 2012-03-20 tsd/TSD2012/paper.tex
tsd paper 9 pages

\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{subfig}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

@@

\begin{abstract}
The work describes a wide-coverage computational grammar for Swedish. 
It is developed using GF(Grammatical Framework), a dependently-typed 
functional language, specialized for grammar programming. We trained
and evaluated the grammar by using Talbanken, the largest treebank for 
Swedish. As a result 65 $\%$ of the Talbanken trees were translated
into the GF format in the training stage. Moreover, we obtained a language
model for Swedish which we use for disambiguation when parsing free text
with the grammar. 
\keywords{grammar-formalism, computational grammar, treebank}
\end{abstract}

@@

\textbf{Ramona: Copied from the thesis, could be improved!}
GF has strong support for multilinguality and has so far been used successfully for 
controlled languages \cite{cnl}, while recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that a parser or grammar is not freely available
does not only restrict its usage but also its possibilities of being further
developed. Our goal is
to create an open-source Swedish grammar from which we derive a parser
accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
\noindent To build the grammar from scratch would be not only time consuming but
would also mean that already existing systems would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
sources.
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules for morphology and syntax.
This is what is meant by grammar in this thesis, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get a well-defined system for describing language, as well as a
strong connection to and possibility of translation between the 
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.

Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
into the GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in the GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism.
It is based on Martin-Löf type theory which allows
reasoning within the programming language.

Our goal is a wide-coverage grammar for Swedish which -- combined with statistical 
resources -- can be used for parsing open-domain text. The grammar implementation gives
a deep analysis, in the trade-off between coverage and depth we aim for depth
while getting coverage from statistical methods.\\

Outline ...

@@

@@

% Chapter 2.4(with emphasis on the syntactic structures)
\textbf{Too long?}
Swedish is a North-Germanic language, sharing most of its grammatical structure 
with Norwegian and Danish. It is also one of the official languages in Finland
and is spoken by approximately 9 million people.
Although Swedish syntax often is similar to English, the  morphology is richer and the
word order more intricate.
It is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but fronting other constituents
-- topicalisation -- is very common,  especially for temporal and
locative adverbial phrases \cite[\textsection 1027]{H&H}.
Fronting the finite verb marks questions.

Swedish has two ways of forming passive verb phrases: the 
\textbf{periphrastic passive}, formed by using the modal auxiliary verb `bli' (\emph{`become'})
and the \textbf{s-passive} which is formed by adding an \emph{s} to the verb: \\
\enumsentence{Passive \hspace{45mm} Active\\
\shortexm{12}
{Tjuven & togs&  av & polisen}
{\emph{the}\emph{ thief}&  \emph{took}+\textbf{s}&  \emph{by}&\emph{the police}}
{``The thief was arrested by the police"}
{&&Polisen & tog & tjuven}
{&&\emph{the police} & \emph{took} & \emph{the thief}}
{\hspace{-9mm}``The police arrested the thief"}}
\label{gfPass:s-pass}

Special reflexive pronouns and reflexive possessive pronouns for the 3rd
exist \cite[\textsection 310 \& 319]{H&H}, distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{llll}
a. & Han slog \textbf{sig}.&\hspace{20mm}b. & Han såg \textbf{sitt} barn.\\
&\emph{He hit \sc self.} && \emph{He saw {\sc self's} child.}
\end{tabular}}
\enumsentence{
\begin{tabular}{llll}
a. &Han slog \textbf{honom}. &b.& Han såg \textbf{hans} barn.\\
&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
\end{tabular}}

\subsection{GF}
\label{sec:gf}
\textbf{Ramona: this is copied too, probably much room for improvement,
and it also has to be shortened}
%Chapter 2.1 + reference to Krasimir's work on the Penn Treebank
%(+ 2.1.3), nothing about resource library here
The main component of the project is the Grammatical Framework (GF) \cite{gfbok},
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and represents a formalism
stronger than mildly context-free grammars.

GF is a strongly typed functional programming language, inspired by
ML \\\cite{ml} and Haskell \cite{haskell}. It is also a logical framework,
and the built-in functionality for logic and reasoning 
is inspired by $\lambda$ Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented at Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. One of the biggest
achievements is a library covering the 
basic morphological and syntactic structures of more than
20 languages (see section \ref{sec:resources}).

A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by Java, Java script, C and Haskell libraries. 
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The shell can be tried out online 
together with an interactive GF editor 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}
Figure \ref{fig:shell} shows commands for parsing, where the results are shown as
in figure \ref{fig:gftree1}. 
GF uses Graphviz\footnote{http://www.graphviz.org/}
to visualize the trees and the given commands (fig. \ref{fig:shell}) specify that the
output format should be PDF and that these files should be opened by the program
Evince. The user can choose to see the \textit{parse tree} (fig. \ref{pic:gfStree})
or the \textit{abstract tree} (fig. \ref{pic:gfAtree}). Abstract trees are more verbose and show all
functions and types used for parsing the sentence. 
\begin{figure}[!h]
\centering
\subfloat[Swedish parse tree]
{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTreec.pdf}}
\label{fig:gftree1}
\subfloat[English parse tree]
{\includegraphics[width=30mm]{gfETreec.pdf}
\label{pic:gfEtree}}
\subfloat[GF abstract tree]
{\label{pic:gfAtree}\includegraphics[width=30mm]{gfTree.pdf}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
Parse trees on the other hand show only the types assigned to the words and phrases.
Information about tense, polarity etc., which are explicitly given in the abstract tree, are
not reproduced in the parse tree. Hence, parse trees do not give complete representations
but model the parse results in a transparent manner.

The scope of GF grammars has so far been controlled language, a subset of a
natural language used for a restricted domain. By
restricting the coverage, the number of ambiguities
can be limited and it can be ensured that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily.

The use of controlled language thus gives the possibility of good and reliable
translation and GF has successfully been used for this in a number of projects.
WebAlt \cite{webalt} aims to develop language independent
material for mathematics problems by using GF grammars.
The formal software verification tool
KeY \cite{key} used GF for translating formal specification to
natural language.
GF has further been used for describing dialogue grammars \cite{talk}, and
the framework is one of the main component in the
European project MOLTO\footnote{http://www.molto-project.eu/}, which develop
online translation between to 15 languages.\\

As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
efficient and easy-going manners are added
and the possibilities to use GF in different environments
increased. There is research on how to make more use of the dependent
types, for reasoning by using ontologies \cite{ontologies2} or generating natural
language via Montague semantics \cite{montague}.

@@

%Chapter 2.2
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
We have used Talbanken \cite{talbanken},
a freely available, manually annotated, large-scale treebank.
It is analyzed with the MAMBA annotation scheme (Teleman, 1974) and 
consists of four
parts. Two of them are transcriptions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.
\\
\noindent Talbanken was also used to train the Swedish version of the Malt parser \cite{malt}
and was then redistributed in an updated version,
Talbanken05 \cite{talbanken05}.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html}
XML-formats
where the trees have been made deeper and more detailed while still containing
the lexical MAMBA layer. 
The Malt parser was trained on section P of Talbanken, and these
more than 6000 sentences have been used our project. 
The treebank has served as an inspiration and an evaluation source throughout the
project.

\section{Related Work}
\textbf{Ramona: also just copied, if you want, improve it}
%Chapter 2.5 - w/o the references about lexical resources
%
%Mention the LREC paper and what it does, without any direct reference
%between the authors
There are many other parsers for Swedish,
like the well-known data-driven Malt parser \cite{malt},
which has been trained on Talbanken. 
There are also a number of grammar-based parsers, although none is freely available.
The cascaded finite state parser CassSwe \cite{casswe} and
The Swedish Constraint Grammar \cite{birn}
give syntactic analyses. 
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
\cite{fdg},
an extension of the Constraint Grammar
formalism, and produces a dependency structure focusing on finding the nominal
arguments.

The LinGO Grammar Matrix \cite{matrix}, is a starter-kit for building Head-Driven Phrase
Structure Grammars \cite{hpsg} (HPSG) providing compatibility with tools for
parsing, evaluation, semantic representations etc.
Translation is supported by using Minimal Recursion
Semantics \cite{mrs} as an interlingua. \\
There is a collection of grammars implemented in this framework, giving broad-coverage
descriptions of
English, Japanese and German.
The Scandinavian Grammar Matrix \cite{scandmatrix} covers common parts of
Scandinavian, while Norsource \cite{norsource} describes Norwegian. A Swedish version
was based upon this (SweCore, Ahrenberg) covering the morphology and some
differences between Swedish and Norwegian. Further, there is the BiTSE 
grammar \cite{stymne}, also implemented using the Lingo Matrix,
which focuses on describing and translating verb frames.

The Swedish version of the Core Language Engine (CLE) \cite{gamback}
gives a full syntactic analysis as well as semantics represented in `Quasi logical form'. A
translation to English  was implemented and the work was further developed in the spoken
language translator \cite{spoken}. Unfortunately, it is no longer available. The coverage of the Swedish

In the TAG formalism \cite{tag}, there are projects on getting open source, wide-coverage grammars
for English and Korean, but, to our knowledge, not for Swedish.

The ParGram \cite{pargram} project aims at making wide coverage grammars using
the Lexical Functional Grammar approach \cite{lfg}.
The grammars are implemented in parallel in order to coordinate the analyses of
different languages and there are now grammars for English, German, Japanese and Norwegian.

@@

%Discussion about the two dimensions of a grammar - lexicon and syntax 
There are to aspects of on a computational grammar clearly distinctive (?)  in
our work- the work on the lexicon and the work regarding syntax.
For this article, we will focus on the syntactical dimension, the description
and formalization of grammar rules and the usage of Talbanken.

@@

%Chapter 2.1.2
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
Since the languages all share the same abstract syntax translation between any given
pair is possible, which is valuable
when implementing multilingual applications.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The grammars cover the morphology,
word order, agreement, tense, basic conjunction etc.
Due to the syntactic similarities of the Scandinavian languages, much of the implementation
is shared with Norwegian and Danish. The modules that concern the lexical
aspects are separate, while 85 \% of the syntax description is shared.
There are about 80 functions, which describe the rules for building phrases and clauses.
\begin{verbatim}
PredVP       : NP -> VP -> Cl ; -- Predication
ComplVPSlash : VPSlash  -> VP ; -- Complementation
\end{verbatim}
The analysis preformed by GF is driven by the parts of speech, which are combined into
parts of sentences.
In addition to the core resource grammars, which is shared with the other languages implemented in the
library, there is also an extra module, simple called the \verb|Extra| module, where
 language specific constructions may bee added
The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.

@@

%Everything about Saldo, Lexin, etc, compressed.
%
%We use the resources from the other paper, so to say :)
\textbf{Some more here}
Our main lexical resource is SALDO \cite{saldo}. We use the tools
described in \cite{ahlbergEnache} to extract a large-scale lexicon and annotate
this with valencies from Lexin. This gives us a dictionary with more than 
100 000 entries, covering all but 500 words from Talbanken ignoring names,
numbers and compounds.

@@

%Section 4.2
%
%Selection of additions to the grammar 
%
%- reflexive possesive pronouns
%
%- s-passive
%
%- impersonal constructions
%
%- miscellaneous 
It has earlier been hard to identify missing constructions of the Swedish
implementation since there was no large resource available to evaluate it on.

Our evaluations are based on Talbanken and when first conducting tests
 we found much room for improvement. The items listed below are examples
 of constructions implemented during this work.

\textbf{shorten these sections}
\subsubsection{The s-passive}
Passive voice is often used in Swedish, especially the
\textit{s-passive} as described in ....
Some studies suggest that the s-passive is used in more than 80 \% of the times
\cite{laanemets}.
It is however not as common in the other Scandinavian languages, 
where not all words have passive forms for all tenses. The resource grammar for
Scandinavian therefore implemented the function for passive,
\verb-PassV2-, by using auxiliary verb.
The function allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.
During this project, the s-passive was added although the
periphrastic passive is still allowed.
The grammar further allows not only transitive verbs, but all verb phrases that
misses an object, to form passives.
A ditransitive verb -- `give' (\ref{sent:give2pass}) -- hence gives rise to two
passives, (\ref{ex:passV32}) and (\ref{ex:passV33}).
\enumsentence{\textbf{Active use of two-place verb}\\\shortex{4}
{Vi & erbjöd& henne& jobbet }
{\emph{we} & \emph{offered}& \emph{her}&\emph{the job}}
{``We offered her the job"}}
\label{sent:give2pass}
\enumsentence{\textbf{First place in two-place verbs}\\\shortex{3}
{Hon & erbjöds & jobbet}
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\textbf{Second place in two-place verbs}\\\shortex{3}
{Jobbet & erbjöds & henne }
{\emph{the job} & \emph{offered+\textbf{s}}&\emph{ her}}
{``The job was offered to her"}}
\label{ex:passV32}

\subsubsection{Impersonal constructions}
\label{sec:Formal}
Formal subjects \cite[\textsection 19]{SAG} are often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{\emph{it }& \emph{sits }& \emph{a} & \emph{bird} &\emph{ on} & \emph{the roof}}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.

Both the verb and the noun phrase have some restrictions: the determiner of the
noun phrase must be of such type that it requires both the noun and its modifiers
to appear in indefinite form. The verb phrase must consists of a intransitive verb
or be in passive form.
A very common example of impersonal constructions is sentences with the verbs \emph{finnas},
(\emph{exist})
\emph{saknas} (\emph{miss}) and \emph{fattas}(\emph{lack}).
\enumsentence{\begin{tabular}[t]{@{}*{18}{l@{\ }}}
a. &Det & finns & kaffe &&&
b. &Det & saknas&  kaffe &&&
c. &Det & fattas & kaffe \\
&\emph{it }& \emph{exist }& \emph{coffee} && &
 &\emph{it }& \emph{misses} & \emph{coffee} && &
 &\emph{it} & \emph{lacks} & \emph{coffee}\\
\end{tabular}\\
{~~''There is coffee"~~~~ ''There is no coffee"~~~~ ''There is no coffee"}}

\subsubsection{Formalizing the rules for reflexive pronouns by using dependent types}
\label{sec:reflexives}
An important area in a Swedish grammar is the treatment of the reflexive pronouns and
the reflexive possessive pronouns.
The reflexives require an antecedent with which they agree in
number and person.
The third person reflexives requires a third person antecedent and
reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in 
sentence (\ref{ex:ReflGenB}) and (\ref{ex:ReflSubj}).
Furthermore, the antecedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).

Still our grammar should accept the sentences (\ref{ex:ReflGen}),
(\ref{sent:rakasig}) and (\ref{ex:ReflAdv}) :

\eenumsentence{
\item
{Sina vantar hade han glömt på tåget.\\
{\sc self's} gloves, he had forgotten on the train.}
\label{ex:ReflGen}
\item
{*Sina vantar var kvar på tåget.\\
{~~ \sc self's} gloves was left on the train.}
\label{ex:ReflGenB}}
\eenumsentence{
\item{
Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
\label{sent:rakasig}
\item
{*Han vill att jag ser på sig.\\
~~He wants me to look at {\sc self}.}\label{sent:sesig}
}
\eenumsentence{
\item{
Han är längre än sin kompis.\\
He is taller than {\sc self's} friend.
}\label{ex:ReflAP} \label{ex:ReflAdv}
\item{*Han och sin kompis läser en bok.\\
He and {\sc self's} friend are reading a book. }
\label{ex:ReflSubj}
}

>>>>

\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{subfig}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{tikz-qtree}

@@

\begin{abstract}
The work describes a wide-coverage computational grammar for Swedish. 
It is developed using GF(Grammatical Framework), a dependently-typed 
functional language, specialized for grammar programming. We trained
and evaluated the grammar by using Talbanken, one of the largest treebanks for 
Swedish. As a result 65 $\%$ of the Talbanken trees were translated
into the GF format in the training stage and 75 \% of the noun phrases
was parsed during the evaluation. Moreover, we obtained a language
model for Swedish which we use for disambiguation when parsing free text
with the grammar. 
\keywords{grammar-formalism, computational grammar, treebank}
\end{abstract}

@@

Two main approaches divide the research in computational linguistics
in the last years. The first one is the wide-coverage method, usually based on
statistics methods, which compensates its usually shallow analysis of the
input with its coverage. The other approach is the opposite - provides
a deep analysis of the data, but it's coverage is limited to
very specialized input. Our work tries to unite the two approaches by
developing a wide-coverage computational grammar for Swedish. We start
with a minimal model of the language and extend its coverage
significantly, without affecting the quality. The grammar was
developed and evaluated against Talbanken\cite{talbanken}, one of the
largest Swedish treebanks.

%The paper is organized as follows: .... \textbf{maybe leave out if no space}

@@

\subsection{GF}
\label{sec:gf}

The computational grammar was developed within GF(Grammatical
Framework) \cite{gfbok}. GF is a dependently-typed grammar formalism,
based on Martin-Löf type theory, which is mainly used for
multilingual natural language applications.

GF grammars are composed of an abstract syntax, interlingua of the
grammar, describing the semantics on a language-independent level and
a number of concrete syntaxes, usually corresponding to natural
languages. These describe how the concepts from the abstract syntax
are described in the language. 
By defining such a grammar, one also obtains a parser for the language
fragment that is described by a concrete grammar and a natural
language generation tool that is able to generate constructions from
the same language fragment. 
In this way, due to the division of the GF
grammars into abstract and concrete, it is possible to achieve 
semantics-preserving translation between any pair of languages from 
the grammar.

However, because the grammar is strict, we can only deal with
constructions that the grammar defines and nothing more. One could
classify the limitations of the natural language grammar in two
categories --- missing lexical items and missing syntactic
structures. For this reasons GF was mainly used so far for controlled
languages\cite{cnl} and, dialogues systems \cite{talk} and interactive 
systems where the user is guided to stay
within the bounds of the grammar by a predictive
parser\cite{parser}, such as a multilingual tourist
phrasebook\cite{phrasebook}. 
However, there is some recent work on parsing
free text --- a grammar for patent claims from the biomedical
domain \cite{patent}. This work highlighted a
number of caveats of real-world text, but in the same time gives
valuable insights that future work on wide-coverage GF grammars can
benefit from.

The current work aims at overcoming the second limitation of GF
grammars --- adding support for the most common syntactic
constructions that could appear in Swedish texts. For some of these we
use dependent types to encode the syntactic phenomena in an elegant
way. The lexicon limitation is not a problem, as we use an existing 
large Swedish lexicon \cite{ahlbergEnache}, which is being continuously
developed and contains more than 120 000 entries.
%... \textbf{Malin, can you say a bit more ?But not more than a couple
%  of lines :)}

@@

% Chapter 2.4(with emphasis on the syntactic structures)
Swedish is a North-Germanic language, sharing most of its grammatical structure 
with Norwegian and Danish. It is also one of the official languages in Finland
and is spoken by approximately 9 million people.
Although Swedish syntax often is similar to English, the  morphology is richer and the
word order more intricate.
It is a verb-second language: %\cite[p.116]{gunlog}: 
the second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but fronting other constituents
-- topicalisation -- is very common,  especially for temporal and
locative adverbial phrases \cite[\textsection 1027]{H&H}.
Fronting the finite verb marks questions.
%\enumsentence{Passive \hspace{45mm} Active\\
%\shortexm{12}
%{Tjuven & togs&  av & polisen}
%{\emph{the}\emph{ thief}&  \emph{took}+\textbf{s}&  \emph{by}&\emph{the police}}
%{``The thief was arrested by the police"}
%{&&Polisen & tog & tjuven}
%{&&\emph{the police} & \emph{took} & \emph{the thief}}
%{\hspace{-9mm}``The police arrested the thief"}}
%\label{gfPass:s-pass}
Special reflexive pronouns and reflexive possessive pronouns for the 3rd
exist \cite[\textsection 310 \& 319]{H&H}, distinct from the normal 3rd person forms.
%\enumsentence{
%\begin{tabular}{llll}
%a. & Han slog \textbf{sig}.&\hspace{6mm}b. & Han såg \textbf{sitt} barn.\\
%&\emph{He hit \sc self.} && \emph{He saw {\sc self's} child.}\\
%c. &Han slog \textbf{honom}. &\hspace{6mm}d.& Han såg \textbf{hans} barn.\\
%&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
%\end{tabular}}

@@

%Chapter 2.2
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
We have used Talbanken \cite{talbanken},
a freely available, manually annotated, large-scale treebank.
It is analyzed with the MAMBA annotation scheme (Teleman, 1974) and 
the more than 6000 sentences we have been using
%consists of four
%parts. Two of them are transcriptions of spoken language, one a collection of
%text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.

Talbanken was also used to train the Swedish version of the Malt parser \cite{malt}
and was then redistributed in an updated version,
Talbanken05 \cite{talbanken05}.
%It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
%and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html}
%XML-formats
%where the trees have been made deeper and more detailed while still containing
%the lexical MAMBA layer. 
%The Malt parser was trained on section P of Talbanken, and these
%more than 6000 sentences have been used our project. 
%The treebank has served as an inspiration and an evaluation source throughout the
%project.

\section{Related Work}
%\textbf{Ramona: also just copied, if you want, improve it}
%Chapter 2.5 - w/o the references about lexical resources
%
%Mention the LREC paper and what it does, without any direct reference
%between the authors
There exist a number of parser for Swedish already, such as the
data-driven Malt parser \cite{malt}, also trained on Talbanken, the 
cascaded finite state parser CassSwe \cite{casswe}, % and
The Swedish Constraint Grammar \cite{birn} and the Swedish FDG, which
uses the Functional Dependency Grammar \cite{fdg}.
%which gives syntactic analyses. \textbf{I don't really know if the
  %last 2 do... could you rephrase please ?}

Among computational grammars for Swedish, we mention the Swedish
version of the Core Language Engine \cite{gamback}, which
provides a comprehensive description of syntax and semantics, as well
as a translation to English. Unfortunately, the resource is not
available for comparison. Other grammars are BiTSE \cite{stymne}, a
Swedish grammar that uses the HPSG format \cite{hpsg} and developed
within the LinGO Matrix library \cite{matrix}.

The current work uses the extensive Swedish
lexicon developed in \cite{ahlbergEnache} and the Swedish resource
grammar as a starting point.

Currently, GF is the leading technology in the European project
MOLTO (Multilingual On-Line Translation)
\footnote{http://www.molto-project.eu/}, which aims at developing
 tools for translating between 15 languages in real-time and with high quality.

Previous examples of larger type-theoretical GF grammars that use 
dependent-types are SUMO-GF \cite{ontologies2}, a GF representation of
SUMO \cite{sumo}, the largest open-source ontology and  a natural
language generation grammar via Montague semantics \cite{montague}.

@@

%Discussion about the two dimensions of a grammar - lexicon and syntax 
There are two aspects of on a computational grammar -- the work on the lexicon
and the work regarding syntax.  For this article, we will focus on the
syntactical dimension of our work, the description and formalization of grammar
rules and the usage of Talbanken.

@@

%Chapter 2.1.2
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%Since the languages all share the same abstract syntax translation between any given
%pair is possible, which is valuable
%when implementing multilingual applications.
The grammars describe how to construct phrases and sentences and how to
decline words. They cover the morphology,
word order, agreement, tense, basic conjunction etc.
Due to the syntactic similarities between the Scandinavian languages, much of
the implementation for Swedish is shared with Norwegian and Danish. The modules
that concern the lexical aspects are separate, while 85 \% of the syntax
description is shared.  There are about 80 functions, which describe the rules
for building phrases and clauses.
\begin{verbatim}
PredVP       : NP -> VP -> Cl ; -- Predication
ComplVPSlash : VPSlash  -> VP ; -- Complementation
\end{verbatim}
%The analysis preformed by GF is driven by the parts of speech, which are combined into
%parts of sentences.
In addition to the core resource grammars, which is shared with all other
languages implemented in the library, there is also an extra module
%, simple called the \verb|Extra| module,
where language specific constructions may bee added
%The functions given here do not have to be translatable
%to all other language, but are meant to cover language specific constructions.

@@

%Everything about Saldo, Lexin, etc, compressed.
%
%We use the resources from the other paper, so to say :)
As mentioned before, our main lexical resource is SALDO \cite{saldo} from
which a GF lexicon has been extracted \cite{ahlbergEnache}.
Valency information, which is a key feature to good parsing using GF,
are extracted from the lexicon
Lexin\footnote{http://spraakbanken.gu.se/lexin/}.
This gives us a dictionary with more than 100 000 entries, covering all but 500
words from Talbanken (excluding names, compounds and numerical expressions).

@@

%Section 4.2
%
%Selection of additions to the grammar 
%
%- reflexive possesive pronouns
%
%- s-passive
%
%- impersonal constructions
%
%- miscellaneous 
It has earlier been hard to identify missing constructions of the Swedish
implementation since there was no large resource available to evaluate it on.
Our evaluations are based on Talbanken and when first conducting tests
 we found much room for improvement. The items listed below are examples
 of constructions implemented during this work.\\

\noindent\textbf{The s-passive}
Swedish has two ways of forming passive verb phrases: the 
periphrastic passive, formed by using the modal auxiliary verb `bli'
(\emph{`become'})
and the s-passive which is formed by adding an \emph{s} to the verb. 
Passive voice is often used, especially the
s-passive, % (sentence \ref{gfPass:s-pass}).
which is used in more than 80 \% of the times according to some studies 
%Some studies suggest that the s-passive is 
\cite{laanemets}.
It is however not as common in the other Scandinavian languages, 
where not all words have passive forms for all tenses. The resource grammar for
Scandinavian therefore only implemented the periphrastic passive.
%\verb-PassV2-, by using auxiliary verb.
%The function allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
%turned into complete verb phrases; they no longer need an object.
During this project, the s-passive was added although the
periphrastic passive is still allowed.
%The grammar further allows not only transitive verbs, but 
%All verb phrases that misses an object are allowed to form passives.
A ditransitive verb -- \emph{ge} (`give') (\ref{sent:give2pass}) -- gives rise to two
passives, (\ref{ex:passV33}) and (\ref{ex:passV32}), both covered by our new implementation.
\eenumsentence{
\item{\textbf{Active use of two-place verb}\\\shortex{4}
{Vi & erbjöd& henne& jobbet }
{\emph{we} & \emph{offered}& \emph{her}&\emph{the job}}
{``We offered her the job"}}
\label{sent:give2pass}
\item{\textbf{Passive use, first place \hspace{3mm}Passive use, second place}\\\shortex{7}
{Hon & erbjöds & jobbet && \hspace{3mm}Jobbet & erbjöds & henne }
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}
&&  \hspace{3mm}\emph{the job} & \emph{offered+\textbf{s}}&\emph{ her}}
{``She was offered the job"}
~``The job was offered to her"}}
\label{ex:passV33}
%\item{\textbf{Second place in two-place verbs}\\\shortex{3}
%{Jobbet & erbjöds & henne }
%{``The job was offered to her"}}
\label{ex:passV32}}

\subsubsection{Impersonal constructions}
\label{sec:Formal}
Formal subjects \cite[\textsection 19]{SAG} are often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{\emph{it }& \emph{sits }& \emph{a} & \emph{bird} &\emph{ on} & \emph{the roof}}
{``There is a bird sitting on the roof''}}
%\emph{`Det'} has the position of the subject, and the real subject, 
%\emph{`en fågel'} the one of an object.
Restrictions on both the verb and the noun phrase are covered by the grammar:
the determiner of the
noun phrase must be of such type that it requires both the noun and its modifiers
to appear in indefinite form. The verb phrase must consist of an intransitive verb
or be in passive form.
%A very common example of impersonal constructions is sentences with the verbs \emph{finnas},
%(\emph{exist})
%\emph{saknas} (\emph{miss}) and \emph{fattas}(\emph{lack}).
%\enumsentence{\begin{tabular}[t]{@{}*{18}{l@{\ }}}
%a. &Det & finns & kaffe &&&
%b. &Det & saknas&  kaffe &&&
%c. &Det & fattas & kaffe \\
%&\emph{it }& \emph{exist }& \emph{coffee} && &
% &\emph{it }& \emph{misses} & \emph{coffee} && &
% &\emph{it} & \emph{lacks} & \emph{coffee}\\
%\end{tabular}\\
%{\indent~~''There is coffee"~~~~ ''There is no coffee"~~~~ ''There is no coffee"}}

\subsubsection{Formalizing the rules for reflexive pronouns by using dependent types}
\label{sec:reflexives}
An important area in a Swedish grammar is the treatment of the reflexive pronouns and
the reflexive possessive pronouns.
The reflexives require an antecedent with which they agree in
number and person. They must 
not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in 
sentence (\ref{ex:ReflGenB}) and (\ref{ex:ReflSubj}).
%Furthermore, the antecedent must be within the same finite sentence as the reflexive
%pronoun, see (\ref{sent:sesig}).
Still our grammar should accept the sentences (\ref{ex:ReflGen})
%(\ref{sent:rakasig}) 
and (\ref{ex:ReflAdv}) :

\eenumsentence{
\item
{Sina vantar hade han glömt på tåget.\\
{\sc self's} gloves, he had forgotten on the train.}
\label{ex:ReflGen}
\item
{*Sina vantar var kvar på tåget.\\
{~~ \sc self's} gloves was left on the train.}
\label{ex:ReflGenB}}
%\eenumsentence{
%\item{
%Jag vill att han rakar sig.\\
%I want him to shave {\sc self}.}
%\label{sent:rakasig}
%\item
%{*Han vill att jag ser på sig.\\
%~~He wants me to look at {\sc self}.}\label{sent:sesig}
%}
\eenumsentence{
\item{
Han är längre än sin kompis.\\
He is taller than {\sc self's} friend.
}\label{ex:ReflAP} \label{ex:ReflAdv}
\item{*Han och sin kompis läser en bok.\\
He and {\sc self's} friend are reading a book. }
\label{ex:ReflSubj}
}

In the standard GF analysis, which is preformed bottom-up starting from the
POS-tags, information about semantic roles are given by the functions, not by the
categories. For example, the first argument of the function 
\verb-PredVP- always acts as the subject, but the noun phrase itself does not carry
information about its semantic role.
As noun phrases containing reflexive pronouns may be used as ordinary
noun phrases, apart from the restrictions stated above,
we do not want them to differentiate
from other \verb-NP- on the type level since we would then need to duplicate
our code.
%due to the rich type system in GF. 
Still, the type system should prevent noun phrases containing reflexive
pronouns from being used as subjects.
This does not only concern noun phrases, but also adverbial and adjectival phrases 
(\ref{ex:ReflAP}).
%\eenumsentence{
%\item{
%taket på sitt hus\\
%\emph{the roof of {\sc self's} house}}
%\item{
%längre än sin kompis\\
%{\emph{taller than {\sc self's} friend}}}}
%\label{ex:adjTyp}
